---
title: "glitter for Wikidata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{recitR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE, message=FALSE, warning=FALSE
)
```

```{r setup}
#if (!require("sf")) install.packages("sf")

if (!require("remotes")) install.packages("remotes")
if(!require("glitter")) remotes::install_github("lvaudor/glitter")
library(glitter)
```


# Find items and properties to build your query

To find the identifiers of items and properties of interest, you can:

- browse [Wikidata](https://www.wikidata.org/wiki/Wikidata:Main_Page) 
- use package `WikidataR` (functions `find_item()`,`find_property()`).

Then `recitR` functions might be used to start exploring data.

# Example 1: Lyon Metro

Imagine you are interested in the Wikidata available regarding the **Lyon Metro network**.

Let's try and see if there are Wikidata about it:

```{r}
find_item("Metro Lyon")
```

So you'd be interested, for instance, in all the subway stations that are part of this network.

```{r}
find_property("part of")
```

So you're looking for all the stations that are part of ("wd:P361") the Lyon metro network ("wd:Q1552").

You could access this information through:

```{r}
stations=get_triplets(t="?items wdt:P361 wd:Q1552")
stations %>% head()
```

Notice that we do not have values yet for the stations (that's what we're looking for) hence the use of "?" at the beginning of the subject string.

To also get the labels for stations, we can use the argument `label`:

```{r}
parts_metro_Lyon=get_triplets(t="?items wdt:P361 wd:Q1552", label="?items")
parts_metro_Lyon %>% head()
```

For now, we get 50 items, not only stations but also other types of items such as metro lines. Let's have a look at the item "Place Guichard - Bourse du Travail" (Q59855) which we know correspond to a station.

The function `get_claims()` of package `recitR` enables you to see all the direct wikidata properties (and their values) based on an item's id. 

```{r}
get_claims("wd:Q599865")
```

Property "wdt:P31" should enable us to collect only stations ("wd:Q928830") instead of all parts of the Lyon Metro network.

We can enrich and refine our query incrementally with `add_triplets()` before sending it (whereas `get_triplets()` was a shortcut to build and send simple requests) :

```{r}
stations_metro_Lyon=add_triplets(t="?stations wdt:P361 wd:Q1552", label="?stations") %>% 
  add_triplets(t="?stations wdt:P31 wd:Q928830") %>% 
  send()
head(stations_metro_Lyon)
```

We now get 42 stations that are part of the Lyon metro network.

If we wanted to get other properties and associated values for these stations (for instance their location ("wdt:P625")) we could proceed this way:

```{r}
stations_metro_Lyon=add_triplets(t="?stations wdt:P361 wd:Q1552", label="?stations") %>% 
  add_triplets(t="?stations wdt:P31 wd:Q928830") %>% 
  add_triplets(t="?stations wdt:P625 ?coords") %>% 
  send()
head(stations_metro_Lyon)
```

This tibble can be easily transform into a Simple feature collection (sfc) :

```{r, eval=FALSE}
stations_metro_Lyon.shp=sf::st_as_sf(stations_metro_Lyon, wkt = "coords")
plot(stations_metro_Lyon.shp$coords)
```


`recitR` provides functions to clean and transform "raw" Wikidata tibbles into tibbles that are easier to use in R.

Function `clean_wikidata_table()` lightens some columns with wikidata-formatted URIs:

```{r}
stations_metro_Lyon=clean_wikidata_table(stations_metro_Lyon)
```


Function `transform_wikidata_coords()` get the coordinates as longitude (`lng`) and latitude (`lat`) based on the Wikidata WKT formatting of spatial coordinates ("Point(*lng* *lat*)").

```{r}
stations_metro_Lyon=stations_metro_Lyon %>% 
  transform_wikidata_coords("coords")
```

The resulting table may then be used easily with (for instance) package `leaflet`:

```{r}
leaflet::leaflet(stations_metro_Lyon) %>%
  leaflet::addTiles() %>%
  leaflet::addCircles(popup=~stationsLabel)
```


# Example 2 : cities around Lyon

Now, let's imagine that we are interested in the cities in a 200km radius around Lyon.

```{r}
find_item("Lyon")
find_item("city")
```

We could start exploring Wikidata with this query which finds all items that are instances ("wdt:P31") of "city" or of any subclass ("wdt:P279") of "city" . This query might return many items so that it seems reasonable to limit the number of items retrieved for now with the argument `limit`

```{r}
add_triplets(t="?city wdt:P31/wdt:P279* wd:Q515", label="?city") %>%
  spq_head(n=10) %>% 
  send()
```
Now, let's get the location ("wdt:P625") of the cities

```{r}
add_triplets(t="?city wdt:P31/wdt:P279* wd:Q515",
             label=c("?city"))
  add_triplets(t="?city wdt:P625 ?coords") %>%
  spq_head(n=10) %>% 
  send()
```

We can refine this query, stating that we want cities (or items of subclasses of city) in a radius of 5km around Lyon (which has lat-long coordinates ~ 45.76 and 4.84). We will use the argument `within_distance`:


```{r}
cities_around_Lyon=add_triplets(t="?city wdt:P31/wdt:P279* wd:Q486972",
                                label="?city") %>% 
  add_triplets(t="?city wdt:P625 ?coords",
               within_distance=list(center=c(long=4.84,lat=45.76),
                                    radius=5)) %>% 
  send()
head(cities_around_Lyon)
```

Actually, rather than getting Lyon's long-lat coordinates "by hand" we could get them directly through our query, which means we have to add a triplet about Lyon ("wd:456") and its coordinates:

```{r, eval=FALSE}
add_triplets(t="wd:Q456 wdt:P625 ?coordLyon") %>% 
  add_triplets(t="?city wdt:P625 ?coords",
               within_distance=list(center="?coordLyon", radius=2)) %>%
  add_triplets(t="?city wdt:P31/wdt:P279* wd:Q486972",
             label="?city") %>%
  build_sparql() %>% cat()
```

