---
title: "glitter for Wikidata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{glitter for Wikidata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE, message=FALSE, warning=FALSE
)
```

```{r setup}
library(glitter)
library(WikidataR)
library(tidyverse)
library(sf)
library(leaflet)
```

This first vignette shows how to use `glitter` to extract data from the **Wikidata SPARQL endpoint**. We imagine here a case study in which one is interested in the **Wikidata items available regarding the Lyon metro network**.

# Find items and properties to build your query

To find the identifiers of items and properties of interest for a particular case study, you can:

- browse [Wikidata](https://www.wikidata.org/wiki/Wikidata:Main_Page)
- use package `WikidataR` (functions `find_item()`,`find_property()`). Here, we will explore that second option

Let's try and find the Wikidata identifier for the Lyon metro network:

```{r}
WikidataR::find_item("Metro Lyon")
```

So you'd be interested, for instance, in all the subway stations that are part of this network. Let's try and find the property identifier that corresponds to this notion:

```{r}
WikidataR::find_property("part of")
```

So you're looking for all the stations that are part of ("wdt:P16") the Lyon metro network ("wd:Q1552").

# Use glitter functions to start exploring data

## Simple single-line query

The `glitter` functions might now be used to start exploring data.

We're looking for items (the "unknown" in our query below, hence the use of a "?") which are part of the Lyon metro network:

```{r single_line}
stations=get_triple("?items wdt:P16 wd:Q1552")
head(stations)
```

To also get the labels for stations, we can use the argument `label`:

```{r single_line_with_label}
parts_metro_Lyon=get_triple("?items wdt:P16 wd:Q1552",
                            label="items")
head(parts_metro_Lyon)
```

## Simple multi-line query

The `get_triple()` function we used above is a good option when you want to send **a very simple query**, yet more often than not one will wish to **gradually complexify** their initial query, so that it's actually preferable to use the **multi-line equivalent** to get_triple, which allows for such increments:

```{r multi_line}
parts_metro_Lyon=spq_init() %>% 
  spq_add("?items wdt:P16 wd:Q1552") %>% 
  spq_label("items") %>% 
  spq_perform()
dim(parts_metro_Lyon)
head(parts_metro_Lyon)
```

## Labelling

The query above, with `spq_label(?items)`, will return a table comprising both `items` (with the Wikidata identifiers) and `itemsLabel` (with the human-readable label corresponding to these items). If the Wikidata unique identifier is not particularly useful, one can use the argument `.overwrite=FALSE` so that only labels will be returned, under the shorter name `items`:

```{r overwrite_labelling}
parts_metro_Lyon=spq_init() %>% 
  spq_add("?items wdt:P16 wd:Q1552") %>% 
  spq_label("items",.overwrite=TRUE) %>% 
  spq_perform()

head(parts_metro_Lyon)
```



# Detail query 

## Add another triple pattern

For now, we get `r nrow(parts_metro_Lyon)` items, which correspond not only to stations but also to other types of items such as metro lines.

Let's have a look at the item "Place Guichard - Bourse du Travail" ("wd:Q599865") which we know correspond to a station. We can do that e.g. through [the Wikidata url associated to this item](https://www.wikidata.org/wiki/Q599865){target="_blank"}.

Hence, the property called "wdt:P31" ("is an instance of") should enable us to collect specifically stations ("wd:Q928830") instead of any part of the Lyon metro network.

```{r stations}
stations=spq_init() %>% 
  spq_add("?station wdt:P16 wd:Q1552") %>% 
  spq_add("?station wdt:P31 wd:Q928830") %>%   # added instruction
  spq_label(station,.overwrite=TRUE) %>% 
  spq_perform()
dim(stations)
head(stations)
```

## Get coordinates

If we want to get the geographical coordinate of these stations (property "wdt:P625") we can proceed this way:

```{r add_coords}
stations_coords=spq_init() %>% 
  spq_add("?station wdt:P16 wd:Q1552") %>% 
  spq_add("?station wdt:P31 wd:Q928830") %>%
  spq_add("?station wdt:P625 ?coords") %>%      # added instruction
  spq_label(station,.overwrite=TRUE) %>% 
  spq_perform()
dim(stations_coords)
head(stations_coords)
```

This tibble can be easily transformed into a Simple feature collection (sfc) object using package `sf` :

```{r stations_as_sf}
stations_sf=st_as_sf(stations_coords, wkt = "coords")
head(stations_sf)
```

The resulting object may then be used easily with (for instance) package `leaflet`:

```{r leaflet_stations}
leaflet(stations_sf) %>%
  addTiles() %>%
  addCircles(popup=~station)
```

# Add property qualifiers

Now, we would like not only to view the stations but also the connecting lines. One property is of particular interest in this prospect: P197, which indicates **which other stations one station is connected to**. To form connecting lines, this information about the connection to other stations need to be complemented by *the involved line* and *direction* of that connection. Hence, we are not only interested in the **values** of the property P197, but also in the **property qualifiers** corresponding to the connecting line (P81) and direction (P5051)

We can thus complete our query this way:

```{r query_prop_qualifiers}
stations_adjacency=spq_init() %>% 
  spq_add("?station wdt:P16 wd:Q1552") %>% 
  spq_add("?station wdt:P31 wd:Q928830") %>%
  spq_add("?station wdt:P625 ?coords") %>%
  spq_add("?station p:P197 ?statement") %>%          # added instruction
  spq_add("?statement ps:P197 ?adjacent") %>%        # added instruction
  spq_add("?statement pq:P81 ?line") %>%             # added instruction
  spq_add("?statement pq:P5051 ?direction") %>%      # added instruction
  spq_label("station", "adjacent", "line", "direction",.overwrite=TRUE) %>% 
  spq_select(-statement) %>% 
  spq_perform() %>% 
  na.omit() %>% 
  select(coords,station,adjacent,line,direction)

head(stations_adjacency)
```

Now, we would like to put the stations **in the right order** so that we will be able to form the connecting lines. 

This **data-wrangling part is a bit tricky** though not directly due to any glitter-related operation).

We define a function `form_line()` which will put the rows in the table of stations in the correct order.

```{r form_line}
form_line=function(adjacencies,direction){
  N=nrow(adjacencies)
  num=rep(NA,N)
  ind=which(adjacencies$adjacent==direction)
  i=N
  num[ind]=i
  while(i>1){
    indnew=which(adjacencies$adjacent==adjacencies$station[ind])
    ind=indnew
    i=i-1
    num[ind]=i
  }
  adjacencies=adjacencies %>% 
    mutate(num=num) %>%
    arrange(num) 
  adjacencies=c(adjacencies$station,
                direction)
  return(adjacencies)
}
```

Now let's **apply this function to all lines and directions possible**:

```{r calc_lines}
stations_lines=stations_adjacency %>% 
  sf::st_drop_geometry() %>% # make this a regular tibble, not sf
  group_by(direction,line) %>% 
  na.omit() %>% 
  tidyr::nest(.key="adj") %>% # have nested "adj" table for each direction-line
  mutate(station=purrr::map(.x=adj,.y=direction,
                            ~form_line(.x,.y))) %>% 
  tidyr::unnest(cols="station") %>% 
  ungroup() %>% 
  left_join(unique(stations_coords), # get corresponding coordinates
            by=c("station")) %>%
  na.omit()
head(stations_lines)
```

`stations_lines` is now an **sf points object** which is properly formatted to be transformed into an **sf lines object** (the stations are in the right order for each line-direction, and the associated coordinates are provided in the table):

```{r stations_lines_sf}
stations_lines_sf=stations_lines %>% 
  sf::st_as_sf(wkt="coords") %>% 
  group_by(direction,line) %>% 
  summarise(do_union=FALSE) %>%   # for each group, and keeping order of points,
  sf::st_cast("LINESTRING")       # form a linestring geometry
stations_lines_sf
```

We can now use this new object to **display the Lyon metro lines on a leaflet map**:

```{r leaflet_lines}
factpal <- colorFactor(topo.colors(8),
                       unique(stations_lines$direction))
leaflet(data=stations_sf) %>%
  addTiles() %>%
  addCircles(popup=~station) %>% 
  addPolylines(data=stations_lines_sf,
               color=~factpal(direction), popup=~direction)
```


