---
title: "How to explore a new base with glitter"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(glitter)
```

Imagine you are tasked with exploring the [Linked Data Service LINDAS provided by the Swiss Federal Archives](https://lindas.admin.ch/sparql#).
You might be able to read examples and docs if present.
But in any case, you could get inspired by the Chapter 11 "SPARQL cookbook" of the ["Learning SPARQL" book by Bob DuCharme](https://www.goodreads.com/book/show/18198885-learning-sparql) to explore the dataset.
Let's go through an example.

## A word of caution

Depending on the dataset you're working with, some queries might just ask _too much_ of the service so proceed with caution.
When in doubt, add a `spq_head()` in your query pipeline, to ask less at a time, or use `spq_count()` to get a sense of how many results there are in total.

## Asking for a subset of all triples

In the code below we'll ask for 10 triples.
Note that we use the `endpoint` argument of `spq_init()` to indicate where to send the query, as well as the `request_type` argument.

How can one know whether a service needs `request_type = "body-form"`?

- The docs might mention it.
- Trial and error.
- In LINDAS' case if you run a request via https://lindas.admin.ch/sparql/ and use your browser's web developer console at the request tab, you can see the request sends the query in the body.

```{r}
library("glitter") 
query_basis = spq_init(
  endpoint = "https://ld.admin.ch/query",
  request_control = spq_control_request(
    request_type = "body-form"
  )
)
query_basis %>%
  spq_add("?s ?p ?o") %>%
  spq_head(n = 10) %>%
  spq_perform() %>%
  knitr::kable()
```

This first query is helpful in that it shows you can do a query!
Its results however can be... more or less helpful.

## Classes

### Find which classes are declared

At this point you might think you need to use some prefixes in your query.
If these prefixes are present in `glitter::usual_prefixes`, you don't need to do anything.
If they're not, use `glitter::spq_prefix()`.

```{r}
query_basis %>%
  spq_add("?class a rdfs:Class") %>%
  spq_head(n = 10) %>%
  spq_perform() %>%
  knitr::kable()
```

How many classes are defined in total?
This query might be too big for the service.

```{r}
query_basis %>%
  spq_add("?class a rdfs:Class") %>%
  spq_count() %>%
  spq_perform()
```

We can do the same query for owl classes instead.

```{r}
query_basis %>%
  spq_add("?class a rdfs:Class") %>%
  spq_head(n = 10) %>%
  spq_perform() %>%
  knitr::kable()
```

Until now we could still be very in the dark as to what the service provides.

### Which classes have instances?

```{r}
query_basis %>%
  spq_add("?instance a ?class") %>%
  spq_arrange(class) %>%
  spq_head(n = 10) %>%
  spq_select(- instance) %>%
  spq_select(class, .spq_duplicate = "distinct") %>%
  spq_perform() %>%
  knitr::kable()
```

### Which classes have the most instances?

```{r}
query_basis %>%
  spq_add("?instance a ?class") %>%
  spq_select(class, .spq_duplicate = "distinct")  %>%
  spq_count(class, sort = TRUE) %>%
  spq_head(20) %>%
  spq_perform() %>%
  knitr::kable()
```

In this case the class names are quite self explanatory but if they were not we could use

```{r}
query_basis %>%
  spq_add("?instance a ?class") %>%
  spq_select(class, .spq_duplicate = "distinct")  %>%
  spq_label(class) %>%
  spq_count(class, class_label, sort = TRUE) %>%
  spq_head(20) %>%
  spq_perform() %>%
  knitr::kable()
```

## Properties

### Find which properties are declared

Note that you could instead use `spq_add("?property a rdfs:Property")` but in this case it returned nothing.

```{r}
query_basis %>%
  spq_add("?property a owl:DatatypeProperty") %>%
  spq_head(n = 10) %>%
  spq_perform() %>%
  knitr::kable()
```
How many properties are defined in total?
This query might be too big for the service.

```{r}
query_basis %>%
  spq_add("?property a owl:DatatypeProperty") %>%
  spq_count() %>%
  spq_perform()
```

### What properties are used?

```{r}
query_basis %>%
  spq_add("?s ?property ?o")  %>%
  spq_select(- s, - o) %>%
  spq_select(property, .spq_duplicate = "distinct") %>%
  spq_head(10) %>%
  spq_perform() %>%
  knitr::kable()
```

### What values does a given property have? 

```{r}
query_basis  %>%
  spq_prefix(prefixes = c("schema" = "http://schema.org/"))%>%
  spq_add("?s schema:addressRegion ?value") %>%
  spq_count(value, sort = TRUE) %>%
  spq_head(10) %>%
  spq_perform() %>%
  knitr::kable()

```

## Which class use a particular property?

One of the properties is `	
https://gont.ch/longName`. 
Which class uses it?

```{r}
query_basis %>%
  spq_prefix(prefixes = c("gont" = "https://gont.ch/")) %>%
  spq_add("?s gont:longName ?o") %>%
  spq_add("?s a ?class") %>%
  spq_select(-o, -s) %>%
  spq_select(class, .spq_duplicate = "distinct") %>%
  spq_head(10) %>%
  spq_perform() %>%
  knitr::kable()
```

## What data is stored about a class's instance?

For each organization, what data is there?

```{r}
query_basis %>%
  spq_prefix(prefixes = c("schema" = "http://schema.org/")) %>%
  spq_add("?s a schema:Organization") %>%
  spq_add("?s ?property ?value") %>%
  spq_select(-value, -s, class, .spq_duplicate = "distinct") %>%
  spq_perform() %>%
  knitr::kable()
```

And for each postal address?

```{r}
query_basis %>%
  spq_prefix(prefixes = c("schema" = "http://schema.org/")) %>%
  spq_add("?s a schema:PostalAddress") %>%
  spq_add("?s ?property ?value") %>%
  spq_select(-value, -s, class, .spq_duplicate = "distinct") %>%
  spq_perform() %>%
  knitr::kable()
```

## Which data or property name includes a certain substring?

```{r}
query_basis %>%
  spq_add("?s ?p ?o") %>%
  spq_filter(str_detect(o, "[Hh]ydro")) %>%
  spq_select(-s, .spq_duplicate = "distinct") %>%
  spq_head(10) %>%
  spq_perform() %>%
  knitr::kable()
```

## An example query based on what we now know



```{r}
query_basis %>%
  spq_prefix(prefixes = c("schema" = "http://schema.org/")) %>%
  spq_add("?s a schema:Organization") %>%
  spq_add("?s schema:name ?name") %>%
  spq_filter(str_detect(name, "swiss")) %>%
  spq_head(10) %>%
  spq_perform() %>%
  knitr::kable()

```
