---
title: "glitter design and internals"
---

This article describes how glitter works, and why.
At this stage of glitter history, feedback and feature requests are most welcome!

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(glitter)
```

The glitter package helps writing SPARQL queries by implementing an internal _domain-specific language_: with glitter, you write code that mostly looks like R code, and end up with a SPARQL query.
For instance:

```{r}
library("glitter")
query <- spq_init() %>%
  spq_add("?item wdt:P31 wd:Q13442814") %>%
  spq_add("?item rdfs:label ?itemTitle") %>%
  spq_filter(str_detect(str_to_lower(itemTitle), 'wikidata')) %>%
  spq_filter(lang(itemTitle) == "en") %>%
  spq_head(n = 5)

query
```

The R code should therefore be easier to write, and read.
The function names and syntax are meant to remind of the tidyverse, and of base R.

Code using glitter will feature:

- a call to `spq_init()`;
- most often, calls to `spq_add()` to add SPARQL triple patterns to the query;
- potentially, calls to `spq_set()` to define helper values like `spq_set(species = c('wd:Q144','wd:Q146', 'wd:Q780'), mayorcode = "wd:Q30185")`;
- potentially, calls to `spq_filter()`, `spq_select()` to filter results;
- potentially, calls to `spq_arrange()`, `spq_head()`, `spq_offset()` to order and trim results;
- lastly a call to `spq_perform()` to send the query and return the results.

# glitter query object

The query object is a list with elements corresponding to e.g. the FILTER part.
Later we might make it an actual class, maybe an R6 one?

It is built by the different calls to `spq_` functions.
The SPARQL query string is assembled by `spq_assemble()`.
Later we might add some linting at that stage.

# glitter tooling

Under the hood, glitter uses

* the [rlang package by Lionel Henry and Hadley Wickham](https://rlang.r-lib.org/) to encapsulate `...` arguments into quosures before handling them. Useful references were [the Metaprogramming section of the Advanced R book by Hadley Wickham](https://adv-r.hadley.nz/metaprogramming.html) as well as the documentation of the [rlang package](https://rlang.r-lib.org/).
* the [xmlparsedata package by Gábor Csárdi](https://r-lib.github.io/xmlparsedata/) to parse the R looking code, and the [xml2 package by Hadley Wickham, Jim Hester and Jeroen Ooms](https://xml2.r-lib.org/), to transform that code into SPARQL code with XPath. Useful references were the documentation of these two packages as well as searching for XPath query examples via a search engine.
* the [httr2 package by Hadley Wickham](https://httr2.r-lib.org/) package to perform queries. We learnt sending such queries thanks to the source code of the [WikidataQueryServiceR package](https://github.com/wikimedia/WikidataQueryServiceR) by Mikhail Popov.

More details in the next sections.

# spq_add()

`spq_add()` works differently from the other `spq_` functions because it looks closer to SPARQL.

Clearly something like `spq_add(query, "?item wdt:P31 wd:Q13442814")` does not look like R code.
The motivation for this is:

- easier copy-pasting from SPARQL query examples;
- the ability to keep the SPARQL concepts.

Triple patterns are parsed by `decompose_triples` that uses string manipulation.

Now, if one wants to go full DSL, it is possible, via `spq_filter()` and `spq_mutate()`.

The triple pattern in `spq_add(query, "?item wdt:P31 wd:Q13442814")` means finding items that are an instance of ("wdt:P31") of a scholarly article ("wd:Q13442814").
With glitter, you can also write it

```r
spq_init() %>%
  spq_filter(item == wdt::P31(wd::Q13442814))
```

This looks more like a normal tidyverse pipeline. Note that the namespacing here is done the R way i.e. `wdt::P31` as opposed to "wdt:P31".

Similary,

```r
spq_init() %>%
  spq_add("wd:Q331676 wdt:P1843 ?statement")
```

adds a variable that is "wdt:P1843" of Sonchus oleraceus ("wd:Q331676"). 
It can be written:

```r
spq_init() %>%
  spq_mutate(statement = wdt::P1843(wd::Q331676))
```

# Other `spq_` functions

The other `spq_` functions `spq_arrange()`, `spq_select()`, `spq_mutate()`, `spq_mutate()` , `spq_filter()`, `spq_summarize()` are the core of the DSL.

They have `...` as arguments where three different things can be passed:

- R-looking snippets (the DSL), for instance `spq_filter(query, lang(itemTitle) == "en")`;
- R-looking snippets as string (for programmatic use), for instance `spq_filter(query, 'lang(itemTitle) == "en"')`;
- SPARQL snippets escaped with `spq()` (for copy-pasting from SPARQL examples), for instance `spq_filter(query, spq('lang(?itemTitle)="en"'))`.

The names of their other arguments starts with a dot to prevent name clashes.

How do we differentiate these three things that users can pass?

* First, arguments are transformed into quosures with `rlang::enquos(...)`.
* Each of them is passed to a function like `spq_treat_argument()`. "Like" as the more complex behavior of `spq_filter()` and `spq_mutate()`, that can accept R-looking snippets that will be translated to either triple patterns or not, warrants a bit more logic.
* In `spq_treat_argument()` we try evaluating the argument via `rlang::eval_tidy()`. 
    * If the result is of the class `spq`, then it means we can use the string as is, it was SPARQL.
    * If the result is another character, it is the code to be translated (example: `'lang(itemTitle) == "en"'`).
    * If it is not, we need to get the text of the expression via `rlang::expr_text(arg) %>% stringr::str_replace("^~", "")` (example: the user wrote `lang(itemTitle) == "en"`).
* In the two latter cases, we have R-looking code to translate into a SPARQL snippet. We first parse the code with xmlparsedata and xml2. Then a series of transformations supported by XPath happen. The glitter package contains a `tibble` called `all_correspondences`:

```{r}
head(glitter::all_correspondences)
```

So all instances of `n(blabla)` become `COUNT(blabla)`. 
We also transform argument names.
Look at the "SELECT" statement below, the `str_c()` function becomes `GROUP_CONCAT()` and its argument `SEPARATOR`. 
Also note that the argument comes after a colon, not a comma like in R. 

```{r}
spq_init() %>%
  spq_summarise(authors = str_c(name, sep = ', '))
```

Later, we need to document these correspondences better, and we need to stress test the DSL with more cases using arguments.

### Special case of `spq_filter()` and `spq_mutate()`

`spq_filter()` receives R-looking fragments that are translated into SPARQL snippets for FILTER... or triple patterns.
`spq_mutate()` receives R-looking fragments that are translated into SPARQL snippets for SELECT... or triple patterns.

At the moment the detection of which is which is based on `::`: if the R-looking fragment contains `::`, we assume it will become a triple pattern.
Later, we need to make this more robust as the function `spq_set()` makes it easier to create synonyms for any subject/verb/object via SPARQL VALUES.

When we assume `spq_filter()`/`spq_mutate()` has received an R-looking fragment meant to be translated to a triple pattern, it is parsed so, not forgetting the order is not the same in the two cases:

* `spq_mutate(object = verb(subject))`;
* `spq_filter(subject == verb(object))`.

# R CMD check hack

The examples using something like

```r
spq_init() %>%
  spq_filter(item == wdt::P31(wd::Q13442814))
```

got flagged as if wdt were a dependency to be stated.
This is understandable.
To bypass it these examples are not examples, they are R chunks in a section called "Some examples".
This means they aren't checked.
Thankfully we have similar code in the real tests!

# Future work

The issue tracker of glitter is quite representative of future work, as well as all sentences starting with "Later" in this article.
As stated at the very beginning of this article, your ideas and comments are welcome.
